Analysis of Programming Students' Prompts with Generative AI

### 2024-09-26 CSEDU2025/Paper.tex
update paper

\title{Analysis of Programming Students' Prompts to Identify Interactions
Patterns with Generative AI}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814} and
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Federal Institute of Education,
Science and Technology, Porto Alegre, Brazil}
\affiliation{\sup{2}Department Telematics Engineering, Carlos III University,
Madrid, Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario,cdk\}@it.uc3m.es}
}

@@

%What is the problem to be solved?

@@

% Are there any existing solutions?

@@

% Which is the best?

@@

% What is its main limitation?

@@

% What do you hope to achieve?

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT 3.5 based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.

\section{\uppercase{Related Work}}

The advancements in GenAI and its use in programming courses have raised
possibilities and concerns among educators and researchers. The study of
\cite{Yilmaz23} found that the main benefit of using ChatGPT in programming
learning is to provide providing fast and mostly correct answers to questions,
improving thinking skills, facilitating debugging, and increasing
self-confidence. In contrast, the study the same appoint limitations such ask
laziness, being unable to answer some questions, or giving incomplete/incorrect
answers, causing professional anxiety in students.

From a cognitive perspective, \cite{Lo24} shown that students were able to
learn from ChatGPT, which increased their understanding and achievement.
However, concerns were raised that the growing use of AI tools might lead to a
decline in critical thinking among students. Teachers and stakeholders should
continue to investigate pedagogical approaches that leverage ChatGPT to enhance
students' understanding and critical thinking. For example, teachers can
instruct students to fact-check and validate information produced by ChatGPT.

Examined students from different demographic groups using AI tools like
ChatGPT, comparing perceptions and impacts among these groups. The results
showed that AI tools had a positive impact on academic performance, especially
for students with less prior experience.

@@

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.

% \cite{wu24} found that ChatGPT-supported learning not only promoted students'
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.5} % Increase the spacing between rows
    \begin{tabular}{|p{3,5cm}|p{11cm}|}
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      \hline
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple-question exercise (Quiz). \\
      \hline
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

The use of GenAI tools can help students learning by providing rapid assistance
and feedback. However, it is important to teach students how to use these tools
properly to promote learning. In this sense, it is important that students are
aware of the cognitive processes.

Metacognition is defined as the awareness about the cognitive processes.
\cite{flavell79} defined a model of cognitive monitoring in four classes of
phenomena that can act or interact with each other, which are: Metacognitive
Knowledge, Metacognitive Experience, Metacognitive Task, and Metacognitive
Strategies. Metacognitive knowledge refers beliefs about what factor or
variables affect the course and outcome of cognitive enterprise. It is
subdivided into three majors categories or variables: Person, Task, and
Strategy. Metacognitive experience refers to the feelings and emotions
associated with cognitive activities. Metacognitive task refers to the type of
task or problem that the individual is facing. Metacognitive strategies refer to
the methods or techniques that individuals use to control their cognitive
processes.

@@

CharlieBot is a ChatGPT 3.5 based bot that provides assistance to students in a
Java programming courses at UC3M. The bot uses the content of UC3M Java
programming course at EdX to supply a RAG(Retrieval-Augmented Generation) system
and thus try to respond more accurately to students' questions \cite{Sun24}. The
main goal of CharlieBot is support students in their learning process by
providing personalized assistance and feedback.

\section{\uppercase{Method}}

The study was conducted in three phases: data collection, categorization, and
analysis. During the first semester of 2024, student interactions with
CharlieBot were collected as part of the Java programming course at UC3M.
The categorization phase involved classifying student messages into eight
categories. The table \ref{tab:categories} shows the categories of analysis and
their description. The initial categories were based on the work of
\cite{Ghimire24}, however, four new categories were added due to the need to
give more detail to the classification. The analysis phase involved to
understand the way that the students interact with CharlieBot.
\section{\uppercase{Results and Discussion}}

@@

@@

% Uncategorized

% Multiple Choice
\subsubsection*{Sequence of Multiple Choice}

Around 16.9\% of the messages were classified as Multiple Choice. These prompts
typically involve students sending one or more exercises to CharlieBot, asking
for the solution.

We identified a sequence of Multiple Choice patterns in some of the
conversations. From the point of view of Metacognitive Knowledge, a possible
interpretation of this complex phenomenon is that students may believe in their
inability to answer several questions. They may also assume that the questions
share a common theme, and solving them together may help consolidate knowledge.
Additionally, they might understand that obtaining the answers quickly is more
efficient.

Regarding Metacognitive Experiences, students may experience feelings such as
cognitive overload, a sense of urgency, and/or confidence in ready-made answers,
which can often lead to relief and a reduction of frustration.

Concerning the Metacognitive about the Task, one interpretation may be that the
student simply wants to complete the task quickly to alleviate the workload,
and/or obtain a correct answer without focusing on deeper learning at
this moment.

In relation to Metacognitive Strategies, students are likely focusing on saving
time and/or possibly lacking planning skills. They may also perceive that
solving the questions on their own requires too much cognitive effort, but this
pattern might reveal a lack of progress monitoring and understanding of
concepts. After obtaining the answers, the metacognitive evaluation process
could lead to feelings of satisfaction, but further reflection may indicate that
the strategy was not effective in supporting deep learning. Depending on this
evaluation, it may influence future decision-making to change the strategy.

% Debugging help
\subsubsection*{Sequence of Debugging Help}

Around 30.9\% of the messages were classified as Debugging Help. Besides of
that, we identified a pattern of a sequences of Debugging Help in the
conversations. These prompts typically start with a message requesting a
Complete Solution or a Code Snippet. After that, the students normally chance
from passive to more active learning strategy.

In terms of Metacognitive Knowledge, the student probably recognizes that they
do not have the necessary tools or knowledge to debug the code on their own.
Concerning the understanding of the task, they may know that debugging is not
just a matter of superficial corrections but requires understanding the
underlying logic. Thus, the student apparently believes that the sequence of
questions related to debugging will help them understand the codes.

In terms of Metacognitive Experiences, some feelings may arise at this moment,
such as frustration or uncertainty in the face of errors, confusion during the
process, and also growing confidence or satisfaction as they better understand
the code and possible corrections.

With respect to Metacognitive related with the objective of the Task, the
student is probably focused on understanding the code in practical terms,
identifying errors, and learning how to better debug in the future.

As for Metacognitive associated with the Strategy, the student has requested
external help and realizes that dividing the debugging task into several
questions can result in a better understanding of the code. They can monitor
their progress throughout the process, adjusting their questions as necessary,
evaluate the effectiveness of the responses received, and reflect on what was
learned during the process.

@@

% Conceptual Questions
\subsubsection*{Sequence of Conceptual Question}

Messages classified as conceptual questions represent around 32\% of all
messages. We have found a pattern of sequences of Conceptual Questions in some
conversations. Some conversations are entirely composed of messages classified
as Conceptual Questions. In other cases, the sequence of Conceptual questions
starts with a request for a Complete Solution or a Code Snippet, followed by
a progression of Conceptual Questions. In both cases, the student is actively
engaging with the bot to understand the concepts and compromised with more
deep learning.

Regarding Metacognitive Knowledge, these sequences of prompts classified as
Conceptual Question, can be interpreted that the students are aware they have
not fully mastered a concept and are trying to understand it, demonstrating a
good level of self-awareness. In this sense, the student realizes that the
problem requires an understanding of concepts in order to advance in their
studies. In this way, the student applies an active learning strategy.

About Metacognitive Experience, several types of feelings may be associated at
this moment, such as actively reflecting on the answers, uncertainty and/or
confusion when trying to understand the concept, frustration for not
progressing, and satisfaction as their questions are answered in a way that
facilitates understanding.

Regarding the type of Metacognition related to objective of the Task, the
students may recognize the difficult and trying to achieve a deeper
understanding of the evolved concepts to later apply in practice.

With respect to Metacognitive Strategies, the student demonstrates a sence of
planning to identify their knowledge gaps. They may start with more general
questions to obtain an overview of the concept and then ask more specific
questions as they gain more clarity. Because the questions are in sequence, the
student probably can adequately monitor their progress.

\subsection*{Interleaving}

@@

\subsection*{Spacing}

% Spacing
Spacing, a well-established learning strategy \cite{Carvalho20}, was examined
by analyzing the time intervals between users’ interactions with CharlieBot.
Although only 7.4\% of the conversations exhibited message delays exceeding 24
hours, 26.19\% involved response times longer than 60 minutes. These findings
suggest that students typically do not extend conversations across multiple days.
However, the data also reveal that some conversations had one or more pauses
exceeding an hour, a practice known to enhance knowledge retention.

>>>>

\title{Analysis of Programming Students' Prompts with Generative AI}

\author{\authorname{Rodrigo Prestes Machado\sup{1}\orcidAuthor{0000-0003-0428-6387},
Carlos Alario Hoyos\sup{2}\orcidAuthor{0000-0002-3082-0814},
Patricia Callejo Pinardo\sup{2}\orcidAuthor{0000-0001-6124-6213},
Iria Manuela Estévez Ayres\sup{2}\orcidAuthor{0000-0002-1047-5398},
Carlos Delgado Kloos\sup{2}\orcidAuthor{0000-0003-4093-3705}}
\affiliation{\sup{1}Department of Informatics, Federal Institute of Education,
Science and Technology, Porto Alegre, Brazil}
\affiliation{\sup{2}Department of Telematics Engineering, Carlos III University,
Madrid, Spain}
\email{\ rodrigo.prestes@poa.ifrs.edu.br, \{calario,pcallejo, ayres,cdk\}@it.uc3m.es}
}

@@

% %What is the problem to be solved?

@@

% % Are there any existing solutions?

@@

% % Which is the best?

@@

% % What is its main limitation?

@@

% % What do you hope to achieve?

This study aims to analyze the interactions between students and CharlieBot,
an ChatGPT 3.5 based bot developed by UC3M that leverages Retrieval-Augmented
Generation (RAG) to enhance its contextual understanding of Java programming.
We created three research questions to guide our analysis:

RQ1: What was the distribution pattern of the classified messages?

RQ2: Was there a time gap between student submissions? If so, which category
showed the largest spacing between submissions?

RQ3: Do students’ conversations with the bot alternate between practical and
theoretical approaches, thereby promoting interleaving?

\section{\uppercase{Theoretical Framework}}

\subsection{Related Work}

% The advancements in GenAI and its use in programming courses have raised
% possibilities and concerns among educators and researchers. The study of
% \cite{Yilmaz23} found that the main benefit of using ChatGPT in programming
% learning is to provide providing fast and mostly correct answers to questions,
% improving thinking skills, facilitating debugging, and increasing
% self-confidence. In contrast, the study the same appoint limitations such ask
% laziness, being unable to answer some questions, or giving incomplete/incorrect
% answers, causing professional anxiety in students.

% From a cognitive perspective, \cite{Lo24} shown that students were able to
% learn from ChatGPT, which increased their understanding and achievement.
% However, concerns were raised that the growing use of AI tools might lead to a
% decline in critical thinking among students. Teachers and stakeholders should
% continue to investigate pedagogical approaches that leverage ChatGPT to enhance
% students' understanding and critical thinking. For example, teachers can
% instruct students to fact-check and validate information produced by ChatGPT.

% Examined students from different demographic groups using AI tools like
% ChatGPT, comparing perceptions and impacts among these groups. The results
% showed that AI tools had a positive impact on academic performance, especially
% for students with less prior experience.

@@

% Regarding self-regulation, \cite{cai23} noted that ChatGPT allowed students to
% regulate their learning pace.
\begin{table*}[htbp]
  \caption{Categories - adapted from \cite{Ghimire24}}
  \begin{center}
    \renewcommand{\arraystretch}{1.6} % Increase the spacing between rows
    \begin{tabular}{p{3cm} p{12cm}} % Remove vertical bars
      \hline
      \textbf{Category} & \textbf{Description} \\
      \hline
      Debugging Help & Prompts that seek help to identify, fix errors, or understand the provided code snippet. \\
      Conceptual Question & Prompts that are more about understanding concepts than specific code. \\
      Student Correction & Prompts where the student corrects the bot. \\
      Code Snippet & Prompts that ask for a specific part of the code, like a function or a segment. \\
      Complete Solution & Prompts that request an entire solution or a complete code snippet. \\
      Multiple Question & Prompts where the user wants to solve a multiple choice exercise (Quiz). \\
      Language Change & Prompts that request a change of idiom. \\
      Uncategorized & Prompts that do not fit into any of the above categories. \\
      \hline
    \end{tabular}
    \label{tab:categories}
  \end{center}
\end{table*}

@@

The use of generative artificial intelligence tools can support the important
mechanism of learning regulation by students \citep{larsen18}, since these
technologies frequently generate automatic and often accurate responses
\cite{Puryear22}, it becomes essential for students to use these tools more
strategically to promote conscious and active learning.

The Theory of Metacognition addresses awareness of mental processes.
\cite{flavell79} proposed a model of metacognitive monitoring that includes four
interrelated phenomena: Metacognitive Knowledge, Metacognitive Experience,
 Metacognitive Goals, and Metacognitive Actions. These processes do not occur
in isolation, but they influence each other, altering cognitive progress over
time.

Metacognitive knowledge includes beliefs about variables that affect the
outcomes of cognitive activities. It is divided into three types: beliefs about
personal abilities, perceived difficulty in the task, and previously used
strategies. Metacognitive Experience refers to the feelings that arise
before, during, and after cognitive activity, such as frustration, confusion,
and satisfaction, among others. Metacognitive goals are the key to regulating
thought as they relate to the goals the individual seeks to achieve, directly
influencing the actions taken. For example, if a student’s goal is to complete
a task quickly, they may adopt a more passive approach to learning. Lastly,
Metacognitive Actions involve the planning, monitoring, and evaluation of
strategies used to achieve the goals. In terms of planning, students can
determine how to approach a task, such as spacing their study sessions
\citep{Ouhao18, Carvalho20}, interleaving topics \citep{Rivers21}, utilizing
retrieval practice \citep{larsen18}, and other strategies.

\section{\uppercase{Material and Method}}

This section outlines the tools and procedures used in this study. CharlieBot
served as the GenAI tool for educational purposes, and the method was structured
into three distinct phases, each designed to systematically assess its
interaction with students.

@@

CharlieBot is an educational bot built on ChatGPT 3.5 and enhanced with
Retrieval-Augmented Generation (RAG) capabilities to improve the precision of
responses provided to students \cite{Sun24}. The RAG feature was developed using
materials from a MOOC, including text, exercises, video transcripts, and other
resources.

\subsection{Method}

The study was conducted in three phases: data collection,
categorization, and analysis. During the data collection phase, students
enrolled in a second semester Java course at a public university in Spain were
introduced to CharlieBot and allowed to use it freely, without being subjected
to a specific methodology for its use. Data were collected anonymously, ensuring
that the records of interactions could not be linked to any individual student
or their academic performance.

In the categorization phase, the students' messages were manually classified
into eight distinct categories for analysis. Initially, \cite{Ghimire24}
proposed four categories: Debugging Help, Conceptual Question, Code Snippet,
and Complete Solution. However, the data collected indicated the need for
additional categories, resulting in the inclusion of four more: Multiple Choice
(Quiz), Student Corrections, Language Change, and Uncategorized. Table
\ref{tab:categories} presents these categories along with their respective
descriptions.

The analysis phase involved the use of Python scripts in the Jupyter Notebook
environment to extract relevant information from previously classified
data. The datasets and code snippets used in this process are available at: YYY.

\section{\uppercase{Results and Discussion}}

The results of the analysis are presented in this section, addressing the
research questions posed in the introduction.

@@

\subsection{Classification of Messages}

@@

@@

\subsection{Spacing}

% Spacing
Spacing, a well-established learning strategy \citep{Carvalho20}, was examined
by analyzing the time intervals between users’ interactions with CharlieBot.
Although only 7.4\% of the conversations exhibited message delays exceeding 24
hours, 26.19\% involved response times longer than 60 minutes. These findings
suggest that students typically do not extend conversations across multiple days.
However, the data also reveal that some conversations had one or more pauses
exceeding an hour, a practice known to enhance knowledge retention.

\subsection{Interleaving}

@@

% % Conceptual Questions
% \subsubsection*{Sequence of Conceptual Question}

% Messages classified as conceptual questions represent around 32\% of all
% messages. We have found a pattern of sequences of Conceptual Questions in some
% conversations. Some conversations are entirely composed of messages classified
% as Conceptual Questions. In other cases, the sequence of Conceptual questions
% starts with a request for a Complete Solution or a Code Snippet, followed by
% a progression of Conceptual Questions. In both cases, the student is actively
% engaging with the bot to understand the concepts and compromised with more
% deep learning.

% Regarding Metacognitive Knowledge, these sequences of prompts classified as
% Conceptual Question, can be interpreted that the students are aware they have
% not fully mastered a concept and are trying to understand it, demonstrating a
% good level of self-awareness. In this sense, the student realizes that the
% problem requires an understanding of concepts in order to advance in their
% studies. In this way, the student applies an active learning strategy.

% About Metacognitive Experience, several types of feelings may be associated at
% this moment, such as actively reflecting on the answers, uncertainty and/or
% confusion when trying to understand the concept, frustration for not
% progressing, and satisfaction as their questions are answered in a way that
% facilitates understanding.

% Regarding the type of Metacognition related to objective of the Task, the
% students may recognize the difficult and trying to achieve a deeper
% understanding of the evolved concepts to later apply in practice.

% With respect to Metacognitive Strategies, the student demonstrates a sence of
% planning to identify their knowledge gaps. They may start with more general
% questions to obtain an overview of the concept and then ask more specific
% questions as they gain more clarity. Because the questions are in sequence, the
% student probably can adequately monitor their progress.

% % Multiple Choice
% \subsubsection*{Sequence of Multiple Choice}

% Around 16.9\% of the messages were classified as Multiple Choice. These prompts
% typically involve students sending one or more exercises to CharlieBot, asking
% for the solution.

% We identified a sequence of Multiple Choice patterns in some of the
% conversations. From the point of view of Metacognitive Knowledge, a possible
% interpretation of this complex phenomenon is that students may believe in their
% inability to answer several questions. They may also assume that the questions
% share a common theme, and solving them together may help consolidate knowledge.
% Additionally, they might understand that obtaining the answers quickly is more
% efficient.

% Regarding Metacognitive Experiences, students may experience feelings such as
% cognitive overload, a sense of urgency, and/or confidence in ready-made answers,
% which can often lead to relief and a reduction of frustration.

% Concerning the Metacognitive about the Task, one interpretation may be that the
% student simply wants to complete the task quickly to alleviate the workload,
% and/or obtain a correct answer without focusing on deeper learning at
% this moment.

% In relation to Metacognitive Strategies, students are likely focusing on saving
% time and/or possibly lacking planning skills. They may also perceive that
% solving the questions on their own requires too much cognitive effort, but this
% pattern might reveal a lack of progress monitoring and understanding of
% concepts. After obtaining the answers, the metacognitive evaluation process
% could lead to feelings of satisfaction, but further reflection may indicate that
% the strategy was not effective in supporting deep learning. Depending on this
% evaluation, it may influence future decision-making to change the strategy.

% % Debugging help
% \subsubsection*{Sequence of Debugging Help}

% Around 30.9\% of the messages were classified as Debugging Help. Besides of
% that, we identified a pattern of a sequences of Debugging Help in the
% conversations. These prompts typically start with a message requesting a
% Complete Solution or a Code Snippet. After that, the students normally chance
% from passive to more active learning strategy.

% In terms of Metacognitive Knowledge, the student probably recognizes that they
% do not have the necessary tools or knowledge to debug the code on their own.
% Concerning the understanding of the task, they may know that debugging is not
% just a matter of superficial corrections but requires understanding the
% underlying logic. Thus, the student apparently believes that the sequence of
% questions related to debugging will help them understand the codes.

% In terms of Metacognitive Experiences, some feelings may arise at this moment,
% such as frustration or uncertainty in the face of errors, confusion during the
% process, and also growing confidence or satisfaction as they better understand
% the code and possible corrections.

% With respect to Metacognitive related with the objective of the Task, the
% student is probably focused on understanding the code in practical terms,
% identifying errors, and learning how to better debug in the future.

% As for Metacognitive associated with the Strategy, the student has requested
% external help and realizes that dividing the debugging task into several
% questions can result in a better understanding of the code. They can monitor
% their progress throughout the process, adjusting their questions as necessary,
% evaluate the effectiveness of the responses received, and reflect on what was
% learned during the process.
